---
title: "Publication bias and sensitivity in R"
subtitle: "Practical · Wednesday 15:22–15:55 · FRB-CESAB training course"
author: "Damien Beillouin"
format:
  html:
    theme: [cosmo, ../../theme/cesab-doc.scss]
    toc: true
    toc-location: left
    toc-title: "Practical"
    number-sections: true
    embed-resources: true
    html-math-method: mathml
    code-copy: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup-hidden
#| include: false
data_file <- "../../donnees/derivees/intercropping_biodiversity.csv"
```

::: {.callout-tip appearance="simple" icon="false"}
## By the end of this practical, you will be able to

1.  Draw and read a **contour-enhanced funnel plot**.
2.  Test for **small-study effects** and **time-lag bias** in a multilevel model.
3.  Run **leave-one-study-out**, **influence** and **study-quality** sensitivity analyses.
4.  Build a **robustness table** for your paper.
:::

| Time | What | Section |
|------|------|---------|
| 15:22 | Load data, fit the main model, funnel plot (code given) | [1](#setup) · [2](#funnel) |
| 15:27 | Classic vs multilevel Egger, PET-PEESE | [3](#tests) |
| 15:37 | Sensitivity: leave-one-study-out, validity | [4](#sensitivity) |
| 15:47 | Your turn — or the *(optional)* parts: time-lag, Cook, fail-safe N | [6](#turn) · [5](#fsn) |
| 15:55 | **Debrief** with the whole group | [7](#key) |

Sections marked *(optional)* are for those who are ahead.

# Setup {#setup}

Open `meta_analysis_course.Rproj` (the course kit), then `td_bias.R` (exercises with blanks `___`). Same data as in the previous practical.

```{r}
#| label: load
library(dplyr)
library(metafor)
library(ggplot2)

# In the participant pack: data_file <- "intercropping_biodiversity.csv"
ic <- read.csv(data_file)

grp <- "Pests"        # <- in section 6 you will change this word only

dat <- ic |>
  filter(functional_group == grp, !sd_missing, !zero_mean) |>
  escalc(measure = "ROM", m1i = b_mean_t, sd1i = b_sd_t, n1i = b_n_t,
         m2i = b_mean_c, sd2i = b_sd_c, n2i = b_n_c, data = _) |>
  mutate(sei = sqrt(vi),                          # standard error
         se_n = sqrt(1 / b_n_t + 1 / b_n_c),     # precision based on sample size only
         inv_n = 1 / b_n_t + 1 / b_n_c,
         year_c = year - mean(year))              # centred publication year

pct <- function(x) round(100 * (exp(x) - 1), 1)

# the main model (from the previous practical)
m <- rma.mv(yi, vi, random = ~ 1 | study_id / es_id, data = dat,
            test = "t", dfs = "contain")
pct(c(m$b, m$ci.lb, m$ci.ub))
```

# Contour-enhanced funnel plot {#funnel}

**Exercise 1.** Draw the funnel with significance contours centred on **zero** (fill in one number, then spend your time on the interpretation).

```{r}
#| label: ex1-template
#| eval: false
funnel(dat$yi, dat$vi, yaxis = "sei",
       level = c(90, 95, 99), shade = c("white", "gray75", "gray90"),
       refline = ___, legend = TRUE, xlab = "lnRR")
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex1-solution
funnel(dat$yi, dat$vi, yaxis = "sei",
       level = c(90, 95, 99), shade = c("white", "gray75", "gray90"),
       refline = 0, legend = TRUE, xlab = "lnRR")
```
:::

::: {.callout-caution icon="false"}
## Interpret (2 min)
Is the funnel symmetric? On which side, and in which zone (white = non-significant), are the imprecise comparisons?
:::

::: {.callout-note collapse="true" title="Answer"}
The bottom of the funnel (imprecise comparisons) is roughly symmetric around 0, with points in the white (non-significant) zone on both sides: no obvious gap where suppressed non-significant results would be. At the very top, `r sum(dat$sei < 0.02)` comparisons have SE < 0.02 — suspiciously precise (two studies; possibly SEs recorded as SDs). Tiny variances are always worth checking at extraction.
:::

# Testing small-study effects and time-lag bias {#tests}

**Exercise 2.** First the *classic* Egger test, which treats every comparison as independent:

```{r}
#| label: egger-classic
regtest(rma(yi, vi, data = dat))
```

Now the **multilevel** version [@nakagawa2022]: add a precision measure as a moderator of the main model.

::: {.callout-warning}
## The lnRR trap
The SE of a lnRR is computed from the means themselves, which creates an artificial correlation between effect and SE. For lnRR (and SMD), use a precision measure based on **sample size only**: `se_n` $= \sqrt{1/n_T + 1/n_C}$.
:::

**Predict first:** will the small-study effect look stronger with `sei` or with `se_n`?

```{r}
#| label: ex2-template
#| eval: false
pet_se <- rma.mv(yi, vi, mods = ~ sei,  random = ~ 1 | study_id / es_id,
                 data = dat, test = "t", dfs = "contain")
pet    <- rma.mv(yi, vi, mods = ~ ___, random = ~ 1 | study_id / es_id,
                 data = dat, test = "t", dfs = "contain")
round(c(slope_p_SE = pet_se$pval[2], slope_p_n = pet$pval[2]), 3)
robust(pet, cluster = study_id, clubSandwich = TRUE)   # cluster-robust check
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex2-solution
pet_se <- rma.mv(yi, vi, mods = ~ sei,  random = ~ 1 | study_id / es_id,
                 data = dat, test = "t", dfs = "contain")
pet    <- rma.mv(yi, vi, mods = ~ se_n, random = ~ 1 | study_id / es_id,
                 data = dat, test = "t", dfs = "contain")
round(c(slope_p_SE = pet_se$pval[2], slope_p_n = pet$pval[2]), 3)
robust(pet, cluster = study_id, clubSandwich = TRUE)
```

With the SE, the slope looks "significant"; with the sample-size measure (and even more with cluster-robust inference) it is not: **no robust evidence of a small-study effect**. The SE-based result was largely the lnRR artefact.
:::

**PET-PEESE: would the conclusion change if there were bias?** The intercept of `pet` estimates the effect of an infinitely large study (PET); PEESE uses `inv_n` instead. Rule [@stanley2014]: if the PET intercept is significant, report PEESE.

```{r}
#| label: petpeese
peese <- rma.mv(yi, vi, mods = ~ inv_n, random = ~ 1 | study_id / es_id,
                data = dat, test = "t", dfs = "contain")
rbind(main  = pct(c(m$b, m$ci.lb, m$ci.ub)),
      PET   = pct(c(coef(pet)[1],   pet$ci.lb[1],   pet$ci.ub[1])),
      PEESE = pct(c(coef(peese)[1], peese$ci.lb[1], peese$ci.ub[1])))
```

The adjusted estimates are, if anything, stronger reductions: the conclusion does not rest on publication bias. With strong heterogeneity, treat them as sensitivity checks, not as "the true effect".

**Exercise 3** *(optional)*. Time-lag bias: do effects change with publication year?

```{r}
#| label: ex3-template
#| eval: false
tl <- rma.mv(yi, vi, mods = ~ ___, random = ~ 1 | study_id / es_id,
             data = dat, test = "t", dfs = "contain")
tl
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex3-solution
tl <- rma.mv(yi, vi, mods = ~ year_c, random = ~ 1 | study_id / es_id,
             data = dat, test = "t", dfs = "contain")
tl
```

Slope ≈ 0 and far from significant: **no evidence** of a decline effect — but with 28 studies the test has little power.
:::

# Sensitivity analyses {#sensitivity}

**Exercise 4. Leave-one-*study*-out.** The loop is given: complete it so that each refit drops **one whole study**. (≈ 1 min to run.)

```{r}
#| label: ex4-template
#| eval: false
ids <- unique(dat$study_id)
loo <- sapply(ids, function(i) {
  f <- rma.mv(yi, vi, random = ~ 1 | study_id / es_id,
              data = dat[dat$study_id ___ i, ], test = "t", dfs = "contain")
  c(est = f$b[1], ci.lb = f$ci.lb, ci.ub = f$ci.ub)
})
pct(range(loo["est", ]))     # range of the pooled effect
any(loo["ci.ub", ] > 0)      # does any CI cross zero?
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex4-solution
ids <- unique(dat$study_id)
loo <- sapply(ids, function(i) {
  f <- rma.mv(yi, vi, random = ~ 1 | study_id / es_id,
              data = dat[dat$study_id != i, ], test = "t", dfs = "contain")
  c(est = f$b[1], ci.lb = f$ci.lb, ci.ub = f$ci.ub)
})
pct(range(loo["est", ]))
any(loo["ci.ub", ] > 0)
```

We leave out **studies**, not rows: rows of a study are not independent.
:::

**Exercise 5** *(optional)*. **Influence** (Cook's distance, by study) [@viechtbauer2010b]:

```{r}
#| label: ex5-template
#| eval: false
cd <- cooks.distance(m, cluster = dat$___)
sort(round(cd, 2), decreasing = TRUE)[1:5]
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex5-solution
cd <- cooks.distance(m, cluster = dat$study_id)
sort(round(cd, 2), decreasing = TRUE)[1:5]
```

No study stands out from the others (the largest distance is small and close to the next ones): no single study drives the result.
:::

**Exercise 6. Study quality.** How many comparisons could not be appraised (`NA`)? Then refit the model with only the comparisons that meet all validity criteria (`validity_biodiv == 1`, coded by the authors of the database), and test validity as a **moderator** (comparing two CIs by eye is not a test).

```{r}
#| label: ex6-template
#| eval: false
m_val <- rma.mv(yi, vi, random = ~ 1 | study_id / es_id,
                data = filter(dat, validity_biodiv == ___), test = "t", dfs = "contain")
pct(c(m_val$b, m_val$ci.lb, m_val$ci.ub))
```

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: ex6-solution
m_val <- rma.mv(yi, vi, random = ~ 1 | study_id / es_id,
                data = filter(dat, validity_biodiv == 1), test = "t", dfs = "contain")
pct(c(m_val$b, m_val$ci.lb, m_val$ci.ub))
m_val$s.nlevels[1]                  # number of studies left
sum(is.na(dat$validity_biodiv))     # comparisons that could not be appraised

m_vmod <- rma.mv(yi, vi, mods = ~ factor(validity_biodiv), random = ~ 1 | study_id / es_id,
                 data = dat, test = "t", dfs = "contain")
m_vmod$pval[2]                      # does validity change the effect?
```

Same estimate, validity does not moderate the effect for pests. Note that a third of the comparisons could not be appraised: report it.
:::

# *(optional)* Fail-safe N: why not {#fsn}

```{r}
#| label: fsn
fsn(yi, vi, data = dat)
```

::: {.callout-caution icon="false"}
## Discuss
Millions of "missing" studies would be needed to cancel the effect. Does that make you confident? What does this number assume about the missing studies, about heterogeneity and about dependence?
:::

::: {.callout-note collapse="true" title="Answer"}
It assumes missing studies have exactly zero effect, ignores heterogeneity and the dependence between comparisons, and says nothing about the size of any bias. It is almost always "reassuring", hence uninformative: **do not report it** [@nakagawa2022].
:::

::: {.callout-note collapse="true" title="(optional) Trim-and-fill, as a sensitivity analysis only"}
Trim-and-fill cannot handle dependent effect sizes; apply it to **one aggregated effect per study** (here assuming ρ = 0.5 within studies, and a common effect within each study — a strong simplification):

```{r}
#| label: trimfill
agg <- aggregate(dat, cluster = study_id, rho = 0.5)
tf  <- trimfill(rma(yi, vi, data = agg))
tf
```

Here no study is imputed. Report it, if at all, as a sensitivity check — never as "the corrected effect".
:::

# Your turn: natural enemies {#turn}

Go back to section 1, set `grp <- "Natural enemies"`, and rerun **sections 1–3 only** (skip the slow leave-one-out). Is there a small-study effect with `se_n`? Would bias explain the (small, non-significant) increase? Then run Exercise 6: does validity matter here?

::: {.callout-note collapse="true" title="Solution"}
```{r}
#| label: turn-solution
ne <- ic |>
  filter(functional_group == "Natural enemies", !sd_missing, !zero_mean) |>
  escalc(measure = "ROM", m1i = b_mean_t, sd1i = b_sd_t, n1i = b_n_t,
         m2i = b_mean_c, sd2i = b_sd_c, n2i = b_n_c, data = _) |>
  mutate(se_n = sqrt(1 / b_n_t + 1 / b_n_c))
f_ne <- function(mods) rma.mv(yi, vi, mods = mods, random = ~ 1 | study_id / es_id,
                              data = ne, test = "t", dfs = "contain")
m_ne   <- f_ne(~ 1)
pet_ne <- f_ne(~ se_n)
val_ne <- f_ne(~ factor(validity_biodiv))
rbind(main = pct(c(m_ne$b, m_ne$ci.lb, m_ne$ci.ub)),
      PET  = pct(c(coef(pet_ne)[1], pet_ne$ci.lb[1], pet_ne$ci.ub[1])))
c(slope_p = pet_ne$pval[2], validity_p = val_ne$pval[2])
coef(val_ne)
```

The increase of natural enemies is small and not significant; the small-study slope is not significant with `se_n`, and the PET estimate is close to zero: the evidence is **inconclusive**, whatever the bias. But **validity matters**: comparisons meeting all validity criteria show a clearly **larger increase** of natural enemies (about +27%) than the others (about −7%) — poorly designed comparisons may be masking a real benefit. That is a moderator — the topic of the next session.
:::

# Debrief: your robustness table {#key}

Fill it in for pests — this is what reviewers expect:

| Analysis | Estimate [95% CI] | Conclusion changes? |
|---|---|---|
| Main multilevel model | | |
| Only valid comparisons | | |
| Leave-one-study-out (range) | | |
| Small-study effect (`se_n`) / PEESE | | |
| Time-lag (slope) | | |

::: {.callout-tip appearance="simple" icon="false"}
1.  Bias is best **prevented** by the search, not corrected by statistics.
2.  **Asymmetry ≠ publication bias**: read the direction; use contour-enhanced funnels.
3.  With several effect sizes per study, use **multilevel** tests — for lnRR with a **sample-size-based** precision; drop fail-safe N; trim-and-fill only as sensitivity.
4.  Report all sensitivity analyses in a **robustness table**.

**Next session (16:15):** *explaining* heterogeneity — meta-regression. (The multilevel Egger test you just ran was already a meta-regression!)
:::

## Going further in the online notebook {.unnumbered}

-   [Publication bias and sensitivity](https://literaturesynthesis.github.io/notebook/chapters/publication-bias.html)
-   [Data extraction and critical appraisal](https://literaturesynthesis.github.io/notebook/chapters/extraction-appraisal.html)
-   [Reading a meta-analysis critically](https://literaturesynthesis.github.io/notebook/chapters/critical-reading.html)

## References

::: {#refs}
:::
